---
title: "BST 210 Homework 5"
author: "Wenjie Gu"
output:
  pdf_document: default
  html_document: default
---

```{r, include=FALSE}
options(tinytex.verbose = TRUE)
```

```{r}
# import and clean dataset
library(haven)
framingham <- read_dta("framingham.dta")
framingham = na.omit(framingham[c('sex', 'bmi', 'age','agecat','death')])
framingham$sex = framingham$sex -1 
```


### Problem 1

#### 1(a)
```{r}
# fit logistic regression model -  linear
bmi.mortality = glm(death~bmi, family = binomial(), data = framingham)
summary(bmi.mortality)
```

Intercept: The odds of death from any cause is estimated to be $e^{-1.949618} \approx 0.14233$ for individuals with bmi score equal zero. (Not sensible in real life)

Slope: The odds of death from any cause for an individual is estimated to be $e^{0.050932 \approx 1.0523$ times higher for every 1-unit increase in his/her bmi value.

```{r}
# Calculate odds ratio for the effect of a 5-unit change in bmi
odds_ratio = exp(5* 0.050932)
cat("The odds ratio for the effect of a 5-unit change in bmi is",odds_ratio, '\n')
lower_bound = exp(5* (0.050932 - 1.96*0.007686))
upper_bound = exp(5* (0.050932 + 1.96*0.007686))
sprintf("The 95 percent CI of OR for the effect of a 5-unit change in bmi is (%f, %f).", lower_bound, upper_bound)
```

#### 1(b)
```{r}
# fit logistic regression model - linear and quadratic bmi
bmi.quad.mortality = glm(death ~ bmi + I(bmi^2), family = binomial, data = framingham)
summary(bmi.quad.mortality)
```

```{r}
anova(bmi.mortality, bmi.quad.mortality, test = 'Chisq')
```

After including the quadratic term, the linear term becomes insignificant.

Since the Likelihood Ratio Test gives p = 0.6936, we fail to reject the null hypothesis where the linear model is sufficient. Therefore, it is not necessary to include the quadratic term.

```{r}
bmi_range = range(framingham$bmi)
bmi_range
```
```{r}
xweight = seq(0,100,0.01)
yweight1 = predict(bmi.mortality,list(bmi = xweight), type = "response")
yweight2 = predict(bmi.quad.mortality, list(bmi = xweight), type = "response")
plot(framingham$death ~ framingham$bmi, col = "black", pch = 20, xlim = c(0,60), main = "Fitted probability curve")
lines(xweight,yweight1, col = "red")
lines(xweight, yweight2, col = "dodgerblue")
legend("bottomright", c("fitted linear logistic regression curve", "fitted quadratic logistic regression curve") ,col = c("red", "dodgerblue") ,lty = "solid", cex = 0.7)
```

With bmi <20, the model with quadratic bmi term will give a slightly higher mortality probability than the model with only the linear term for a fixed bmi value. For bmi in the approximate range of (20, 35), the two models give similar prediction (the curves overlap). For bmi >35, the model with quadratic bmi term will again give higher prediction result for a fixed bmi than the linear one. However, the trend and shape of the fitted curves is similar.

#### 1(c)
```{r}
odds = function(bmi){
  log_odds = -1.6476413 + 0.0287998 * bmi + 0.0003947 * bmi^2
  odds = exp(log_odds)
  return (odds)
}
oddsratio1 = odds(25)/odds(20)
oddsratio2 = odds(35)/odds(30)
```
```{r}
cat("The odds ratio for a 5-unit increase in BMI (comparing 25 to 20) is:", oddsratio1, '\n')
cat("The odds ratio for a 5-unit increase in BMI (comparing 35 to 30) is:", oddsratio2)
```

####1(d)
```{r}
# Two sample t-test comparing the average bmi of males and females
t.test(bmi~sex, data = framingham)
boxplot(bmi~sex, data = framingham)
```
By performing a t-test comparing the average bmi of males and females, we get a p-value less than 0.05 (p = 1.56e-6). There is evidence suggesting the association between participant sex and bmi.

```{r}
# check if sex is a confounder  
bmi.sex.mortality = glm(death~ bmi + sex, family = binomial, data = framingham)
summary(bmi.sex.mortality)
summary(bmi.mortality)
```

Coefficient for bmi changes from 0.050932 to 0.047439 (% of change: -6.8%), which is less than 10%. Therefore sex is not a confounder.

```{r}
# check if sex is an effect modifier
bmi.sexint.mortality = glm(death~bmi+bmi*sex, family = binomial, data = framingham)
summary(bmi.sexint.mortality)
```

The interaction between bmi and sex has significant coefficient (p = 0.000502), therefore sex is an effect modifier of the effect of continuous BMI on mortality.

For males, the odds of mortality is estimated to be $e^{0.009139} \approx 1.00918$ times higher for every 1-unit increase in bmi value.
For females, the odds of mortality is estimated to be $e^{0.009139 + 0.057502 \approx 1.0689}$ times higher for every 1-unit increase in bmi value.

#### 1(e)
```{r}
age.cont.mortality = glm(death~age, family = binomial, data = framingham)
age.cat.factor.mortality = glm(death~as.factor(agecat), family = binomial, data = framingham)
age.cat.cont.mortality= glm(death~agecat, family = binomial, data = framingham)
```
```{r}
anova(age.cat.cont.mortality,age.cat.factor.mortality, test = "Chisq")

AIC(age.cont.mortality)
AIC(age.cat.cont.mortality)
AIC(age.cat.factor.mortality)

BIC(age.cont.mortality)
BIC(age.cat.cont.mortality)
BIC(age.cat.factor.mortality)

```



```{r}
library(ResourceSelection)
library(LogisticDx)
fitted.cont.results = ifelse(fitted(age.cont.mortality) > 0.5,1,0)
fitted.cat.cont.results = ifelse(fitted(age.cat.cont.mortality) > 0.5,1,0)
fitted.cat.factor.results = ifelse(fitted(age.cat.factor.mortality) > 0.5,1,0)

hoslem.test(framingham$death,fitted(age.cont.mortality) ,g=10)
chisq.test(framingham$death,fitted(age.cat.cont.mortality))
chisq.test(framingham$death,fitted(age.cat.factor.mortality))
```
Of the three models, the continuous age model has the lowest AIC and BIC scores.
After assessing for the goodness of fits of the three models, we found that  all the three models give significant Hosmer-Lemeshow/Pearson Chi-squared statistics, indicating that none of the fits are adequate.
However, if we have to select one "best" model, according to AIC and BIC values, the preferred one should be the continuous age model.

#### 1(f)
```{r}
bmi.agecat.mortality = glm(death~bmi + agecat, family=binomial, data = framingham)
bmi.agecatint.mortality = glm(death ~ bmi + agecat*bmi, family = binomial, data = framingham)
summary(bmi.mortality)
summary(bmi.agecat.mortality)
summary(bmi.agecatint.mortality)
```
The coefficient of bmi changes from 0.050932 to 0.029577 (-36% change). Thus age category is a confounder for the effect of bmi on mortality.
Whereas, the interaction between age category and bmi is not significant, indicating that age category is not an effect modifier for the effect of bmi on mortality.