---
title: "Group Project 2"
author: "Eileen Zhang"
date: "12/3/2019"
output: html_document
---
Data Clean-up
```{r}
#setwd("C:/Users/eilee/Downloads")
framingham <- read_dta(file=file.choose())
dat<-framingham[(framingham$prevap==0 & framingham$prevchd==0 & framingham$prevmi==0 & framingham$prevstrk==0),]
dat$sex<- dat$sex - 1
dat<-dat%>% drop_na(bpmeds)
dat$sex = as.factor(dat$sex)
dat$prevhyp = as.factor(dat$prevhyp)
dat$agecat = as.factor(dat$agecat)
dat$bpmeds = as.factor(dat$bpmeds)
dat$diabetes = as.factor(dat$diabetes)
```

Binary Logistic Model (Effect of prevhyp on death)
```{r}
datsub<-dat[c(1,2,4:12,17,18,35)]
datsub<-na.omit(datsub)
modlogit.full<-glm(death ~.,family=binomial(),data=datsub)
modlogit.step<-step(modlogit.full,trace=0)
summary(modlogit.step)
```

```{r}
modlogit<-glm(death~prevhyp,family=binomial(),data=datsub)
summary(modlogit)
exp(coef(modlogit))
```
OR = 3.041469

```{r}
modlogit.agecat<-glm(death~prevhyp+agecat,family=binomial(),data=datsub)
summary(modlogit.agecat)
modlogit.agecat.emm<-glm(death~prevhyp*agecat,family=binomial(),data=datsub)
summary(modlogit.agecat.emm)
```
Agecat is a confounder for prevhyp but not a EMM.

```{r}
modlogit.sex<-glm(death~prevhyp+sex,family=binomial(),data=datsub)
summary(modlogit.sex)
modlogit.sex.emm<-glm(death~prevhyp*sex,family=binomial(),data=datsub)
summary(modlogit.sex.emm)
```
Sex is not a confounder for prevhype, but a EMM.

```{r}
modlogit.test.interact<-glm(death~ prevhyp + agecat + sex*prevhyp, family = binomial(), 
    data = datsub)
summary(modlogit.test.interact)
modlogit.test<-glm(death~ prevhyp + agecat + sex, family = binomial(), 
    data = datsub)
summary(modlogit.test)
anova(modlogit.test.interact,modlogit.test,test="Chisq")
exp(coef(modlogit.test))
```
OR = 2.1709706

```{r}
modlogit.opt<-glm(death~ sex + sysbp + diabp + cursmoke + cigpday + 
    diabetes + bpmeds + glucose + prevhyp + agecat + sex*prevhyp, family = binomial(), 
    data = datsub)
summary(modlogit.opt)
anova(modlogit.step,modlogit.opt, test="Chisq")
```

```{r}
exp(coef(modlogit.step))
library(ROCR)
library(ResourceSelection)
library(LogisticDx)
fitted.logit_death=predict(modlogit.step,newdata = datsub,type = 'response')
pred <- prediction(fitted.logit_death, datsub$death)
perf <- performance(pred,"tpr","fpr")
plot(perf,colorize=T)
gof(modlogit.step)
```
OR comparing prevhyp =1 to 0, 1.33307692
AUC = 78.7% (77.1%-80.2%)

----------------------------------------------------------------------------------------------------------------------------------------------------
Cox proportional model
```{r}
library(survival)
library(survminer)
dat$survobject = Surv(dat$timedth,dat$death)
datsub.sur<-dat[c(1,2,4:12,17,35,37)]
datsub.sur<-na.omit(datsub.sur)
modsur.full <- coxph(survobject~.,data=datsub.sur)
summary(modsur.full)
```
Stepwise Selection
```{r}
modsur.step <-step(modsur.full,trace=0)
summary(modsur.step)
```
ROC modsur.step
```{r}
library(survivalROC)
nobs <- NROW(datsub.sur)
cutoff <- 1000
## MAYOSCORE 4, METHOD = NNE
Mayo4.0= survivalROC(Stime=datsub3.sur$timedth,  
                     status=datsub3.sur$death,      
                     marker = predict(modsur.step,datsub.sur,type='lp'),     
                     predict.time = cutoff,span = 0.25*nobs^(-0.20) )
plot(Mayo4.0$FP, Mayo4.0$TP, type="l", xlim=c(0,1), ylim=c(0,1),   
     xlab=paste( "FP", "", "AUC = ",round(Mayo4.0$AUC,3)), 
     ylab="TP",main="Mayoscore 4, Method = NNE 
     Year = 1")
abline(0,1)
```
Cutoff > 1000, AUC =0.709

Assessing PH assumption
(1) Using Schoenfeld residuals 
```{r}
test.ph <- cox.zph(modsur.step)
test.ph
```
Add time interaction term for sex
Eliminate insignificant covariates
```{r}
dat$survobject = Surv(dat$timedth,dat$death)
datsub2.sur<-dat[c(1,2,4:12,17,35,37,32)]
datsub2.sur<-na.omit(datsub2.sur)
datsub2.sur$agecat=as.factor(datsub2.sur$agecat)
datsub2.sur$sex=as.factor(datsub2.sur$sex)
datsub2.sur$diabetes=as.factor(datsub2.sur$diabetes)
datsub2.sur$prevhyp=as.factor(datsub2.sur$prevhyp)
datsub2.sur$bpmeds=as.factor(datsub2.sur$bpmeds)
datsub2.sur$sex_time=(as.numeric(datsub2.sur$sex)-1)*log(datsub2.sur$timedth)
modsur.opt=coxph(survobject~sex+sex_time+sysbp+diabp+cigpday+diabetes+bpmeds+glucose+prevhyp+agecat,data = datsub2.sur)
summary(modsur.opt)
modsur.opt2=coxph(survobject~sex+sex_time+sysbp+cigpday+diabetes+prevhyp+agecat,data = datsub2.sur)
summary(modsur.opt2)
```
HR for prehyp = 0.7903

ROC modsur.opt2
```{r}
library(survivalROC)
datsub3.sur<-dat[c(18,32,1,2,4:12,17,35,37,32)]
datsub3.sur<-na.omit(datsub3.sur)
datsub3.sur=datsub3.sur[,c(1,2)]
nobs <- NROW(datsub3.sur)
cutoff <- 1000
## MAYOSCORE 4, METHOD = NNE
Mayo4.1= survivalROC(Stime=datsub3.sur$timedth,  
                     status=datsub3.sur$death,      
                     marker = predict(modsur.opt2,datsub2.sur,type='lp'),     
                     predict.time = cutoff,span = 0.25*nobs^(-0.20) )
plot(Mayo4.1$FP, Mayo4.1$TP, type="l", xlim=c(0,1), ylim=c(0,1),   
     xlab=paste( "FP", "", "AUC = ",round(Mayo4.1$AUC,3)), 
     ylab="TP",main="Mayoscore 4, Method = NNE 
     Year = 1")
abline(0,1)
```
Cutoff > 1000, AUC =0.863

Survival Plot for different agecat using modsur.opt2

Agecat = 1
```{r}
newdata1=data.frame(sex = factor(c(0,0,1,1)), prevhyp = factor(c(0,1,0,1)),
                               agecat = factor(c(1,1,1,1)),diabetes = factor(c(0,0,0,0)),
                               sysbp = rep(mean(datsub2.sur$sysbp, na.rm = TRUE), 4),
                               sex_time = c(rep(mean(datsub2.sur$sex_time[datsub2.sur$sex==0], na.rm = TRUE),2),rep(mean(datsub2.sur$sex_time[datsub2.sur$sex==1], na.rm = TRUE), 2)),
                               cigpday = rep(mean(datsub2.sur$cigpday, na.rm = TRUE), 4))

ggsurvplot(survfit(modsur.opt2,newdata =newdata1),data=datsub2.sur)
```
Agecat = 2
```{r}
newdata2=data.frame(sex = factor(c(0,0,1,1)), prevhyp = factor(c(0,1,0,1)),
                               agecat = factor(c(2,2,2,2)),diabetes = factor(c(0,0,0,0)),
                               sysbp = rep(mean(datsub2.sur$sysbp, na.rm = TRUE), 4),
                               sex_time = c(rep(mean(datsub2.sur$sex_time[datsub2.sur$sex==0], na.rm = TRUE),2),rep(mean(datsub2.sur$sex_time[datsub2.sur$sex==1], na.rm = TRUE), 2)),
                               cigpday = rep(mean(datsub2.sur$cigpday, na.rm = TRUE), 4))

ggsurvplot(survfit(modsur.opt2,newdata =newdata2),data=datsub2.sur)
```
Agecat = 3
```{r}
newdata3=data.frame(sex = factor(c(0,0,1,1)), prevhyp = factor(c(0,1,0,1)),
                               agecat = factor(c(3,3,3,3)),diabetes = factor(c(0,0,0,0)),
                               sysbp = rep(mean(datsub2.sur$sysbp, na.rm = TRUE), 4),
                               sex_time = c(rep(mean(datsub2.sur$sex_time[datsub2.sur$sex==0], na.rm = TRUE),2),rep(mean(datsub2.sur$sex_time[datsub2.sur$sex==1], na.rm = TRUE), 2)),
                               cigpday = rep(mean(datsub2.sur$cigpday, na.rm = TRUE), 4))

ggsurvplot(survfit(modsur.opt2,newdata =newdata3),data=datsub2.sur)
```
Agecat = 4
```{r}
newdata4=data.frame(sex = factor(c(0,0,1,1)), prevhyp = factor(c(0,1,0,1)),
                               agecat = factor(c(4,4,4,4)),diabetes = factor(c(0,0,0,0)),
                               sysbp = rep(mean(datsub2.sur$sysbp, na.rm = TRUE), 4),
                               sex_time = c(rep(mean(datsub2.sur$sex_time[datsub2.sur$sex==0], na.rm = TRUE),2),rep(mean(datsub2.sur$sex_time[datsub2.sur$sex==1], na.rm = TRUE), 2)),
                               cigpday = rep(mean(datsub2.sur$cigpday, na.rm = TRUE), 4))

ggsurvplot(survfit(modsur.opt2,newdata =newdata4),data=datsub2.sur)
``` 

Log-rank test using modsur.opt2
Agecat=1
```{r}
data1=datsub2.sur[datsub2.sur$agecat==1,]
survdiff(survobject~prevhyp+strata(sex),data1)
survdiff(survobject~sex+strata(prevhyp),data1)
```
Control sex, prevhyp significantly different
control prevhyp, sex not significantly different

Agecat=2
```{r}
data2=datsub2.sur[datsub2.sur$agecat==2,]
survdiff(survobject~prevhyp+strata(sex),data2)
survdiff(survobject~sex+strata(prevhyp),data2)
```
Control sex, prevhyp significantly different
control prevhyp, sex significantly different

Agecat=3
```{r}
data3=datsub2.sur[datsub2.sur$agecat==3,]
survdiff(survobject~prevhyp+strata(sex),data3)
survdiff(survobject~sex+strata(prevhyp),data3)
```
Control sex, prevhyp significantly different
control prevhyp, sex significantly different

Agecat=4
```{r}
data4=datsub2.sur[datsub2.sur$agecat==4,]
survdiff(survobject~prevhyp+strata(sex),data4)
survdiff(survobject~sex+strata(prevhyp),data4)
```
Control sex, prevhyp significantly different
control prevhyp, sex significantly different